Disentangling patterns of gene expression in high grade serous ovarian cancer by matrix factorization

### 2020-08-13 MSc_Dissertation_B156476.tex
Added CD38 results

We found no convincing evidence of this, for an individual factor (metasample).  Some components were predictive in one or other of the two datasets, but not both, which is unconvincing.  The factor with GEA showed indicated regulation of angiogenesis (NMF-1), and thence might have prognostic value, performed poorly in survival analysis.   It is possible that a more sophisticated multiple regression model would be successful, but this would risk over-fitting given the limited dataset size.

@@

\noindent
\emph{Which MF method is best suited to transcripomics analysis?}

@@

\section*{Acknowledgements}
I am most grateful to Dr. Ailith Ewing for proposing this project and making available datasets.  Ailith provided excellent, inciteful supervision throughout and gave helpful feedback on an early draft of this dissertation.

I'd like to thank my employer, Canon Medical Research Ltd., for allowing me the flexibility and study leave to undertake this MSc.

Thank you to my family for excusing me from cooking and washing up duties during the more stressful phases of MSc study!

@@

\begin{figure}[htb!]
\begin{center}
\includegraphics[scale=\S]{\plotspath{BOTH_AOCS_TCGA/FactorClustering/multiple_single_factors_scatter_BOTH_NMF_2_3_5_10_bootstrap.pdf}} \\
\includegraphics[scale=\S]{\plotspath{BOTH_AOCS_TCGA/FactorClustering/multiple_single_factors_scatter_BOTH_ICA_2_3_5_10_bootstrap.pdf}} \\
\includegraphics[scale=\S]{\plotspath{BOTH_AOCS_TCGA/FactorClustering/multiple_single_factors_scatter_BOTH_PCA_2_3_5_10_bootstrap.pdf}} \\
%\includegraphics[scale=0.55]{\plotspath{/BOTH_AOCS_TCGA/FactorClustering/silhouette_plots_TCGA_2_3_4_5_6_7_8_9_10_bootstrap.pdf}} 
\caption{Metagene clustering for all three methods applied to the combined N=80+374 AOCS + TCGA dataset with bootstrap sampling, over a range of factorization ranks.  The silhouette scores are also plotted (bottom).  }
\label{fig-BOTH-all-bootstrap}
\end{center}
\end{figure}

>>>>

We found no convincing evidence of this, for an individual factor (metasample).  Some components were predictive in one or other of the two datasets, but not both, which is unconvincing.  The factor which GEA indicated regulation of angiogenesis (NMF-1), and thence might have prognostic value, performed poorly in survival analysis.

Is there a signal for survival to be found in these data?  We have noted that according to \cite{Zhu2020}, the single gene CD38 is highly predictive of survival in EOC.  Does this hold for the TCGA and AOCS datasets we are working with?  This is easy to check, running survival analysis stratified by a hand-crafted metagene, setting the CD38 element to 1, all others to 0.   This was applied as in section \ref{sec-survival-analysis} to TCGA (OS),  AOCS (OS) and AOCS (PFS) -- see figure \ref{fig-cd38 survival}.   The effects are in a consistent direction across the three tests, with high CD38 expression predicting favourable survival.  This is highly significant in the TCGA dataset and $p < 0.05 $ in the smaller AOCS.  So there is a signal, but it did not emerge unbidden from matrix factorization.  (CD38 is in fact highlighted by component ICA-5, see figure \ref{fig-ICA-5-5-go-lineage}, but must be diluted by other genes).

\begin{figure}
\begin{center}
\includegraphics[scale=0.7,draft=false]{\plotspath{SurvivalAnalysis/gene_specific_CD38_survival.pdf}}
\end{center}
\caption{Survival analysis plots stratified by above/below median expression of CD38 only.}
\label{fig-cd38 survival}
\end{figure}

@@

\subsubsection*{Which MF method is best suited to transcripomics analysis?}

@@

\section*{Acknowledgements}
I am most grateful to Dr. Ailith Ewing for proposing this project and making available the datasets.  Ailith provided excellent, insightful supervision throughout and gave valuable feedback on an early draft of this dissertation.  I'd like to thank my employer, Canon Medical Research Ltd., for allowing me the flexibility and study leave to undertake this MSc.  Thank you to my family for excusing me from cooking and washing up duties during the more stressful phases of MSc study!

@@

\subsection{Additional plots relating to the batch effect investigation}
\begin{figure}[htb!]
\begin{center}
\includegraphics[scale=\S]{\plotspath{BOTH_AOCS_TCGA/FactorClustering/multiple_single_factors_scatter_BOTH_NMF_2_3_5_10_bootstrap.pdf}} \\
\includegraphics[scale=\S]{\plotspath{BOTH_AOCS_TCGA/FactorClustering/multiple_single_factors_scatter_BOTH_ICA_2_3_5_10_bootstrap.pdf}} \\
\includegraphics[scale=\S]{\plotspath{BOTH_AOCS_TCGA/FactorClustering/multiple_single_factors_scatter_BOTH_PCA_2_3_5_10_bootstrap.pdf}} \\
\includegraphics[scale=0.55]{\plotspath{/BOTH_AOCS_TCGA/FactorClustering/silhouette_plots_BOTH_2_3_4_5_6_7_8_9_10_bootstrap.pdf}} 
\caption{Metagene clustering for all three methods applied to the combined N=80+374 AOCS + TCGA dataset with bootstrap sampling, over a range of factorization ranks.  The silhouette scores are also plotted (bottom).  }
\label{fig-BOTH-all-bootstrap}
\end{center}
\end{figure}
